---
title: "Week 9"
Subtitle: "Bayesian Statistics"
---


```{r}
#| include: false
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE)
library(tinytex)
library(tidyverse)
```

## Introduction

Suppose a coffee company is testing a new drink. Before adding it to the menu, the company wants to know whether customers genuinely support the launch. Rather than asking for a single opinion, the company collects short customer comments from taste-testing sessions. Some comments clearly support launching the drink. Others are negative, uncertain, or only mildly positive.

Your task is to act as the product team. Each day, you will read a small set of customer comments and decide which ones count as **genuine support** for launching the drink.

For each day:

* Tick the comments that you think count as genuine support.
* The app will count how many comments you selected.
* That count becomes the data for the Bayesian update.

```{=html}
<div class="app-frame-container" id="cafe-app-container">

  <div class="app-frame-toolbar">
    <button type="button" onclick="minimiseCafeApp()" title="Minimise">−</button>
    <button type="button" onclick="maximiseCafeApp()" title="Maximise">□</button>
    <button type="button" onclick="fullscreenCafeApp()" title="Fullscreen">⛶</button>
  </div>

  <iframe
    id="cafe-app-frame"
    src="interactive/Cafe.html"
    width="100%"
    height="620"
    style="border: none;"
    allowfullscreen>
  </iframe>

</div>

<style>
.app-frame-container {
  width: 100%;
  border: 1px solid #d1d5db;
  border-radius: 8px;
  overflow: hidden;
  background: white;
}

.app-frame-toolbar {
  display: flex;
  justify-content: flex-end;
  gap: 6px;
  padding: 6px;
  background: #f3f4f6;
  border-bottom: 1px solid #d1d5db;
}

.app-frame-toolbar button {
  border: 1px solid #9ca3af;
  border-radius: 6px;
  padding: 5px 10px;
  background: white;
  cursor: pointer;
}

.app-frame-container.maximised iframe {
  height: 90vh;
}

.app-frame-container.minimised iframe {
  display: none;
}

.app-frame-container:fullscreen {
  display: flex;
  flex-direction: column;
  width: 100vw;
  height: 100vh;
  border: none;
  border-radius: 0;
}

.app-frame-container:fullscreen iframe {
  flex: 1;
  width: 100%;
  height: 100%;
}
</style>

<script>
function minimiseCafeApp() {
  const container = document.getElementById("cafe-app-container");

  container.classList.toggle("minimised");
  container.classList.remove("maximised");
}

function maximiseCafeApp() {
  const container = document.getElementById("cafe-app-container");

  container.classList.toggle("maximised");
  container.classList.remove("minimised");
}

function fullscreenCafeApp() {
  const container = document.getElementById("cafe-app-container");

  container.classList.remove("minimised");

  if (!document.fullscreenElement) {
    container.requestFullscreen();
  } else {
    document.exitFullscreen();
  }
}
</script>
```

## Tasks

### Part 1: Customer feedback and Bayesian updating

In the app, you acted as the product team for a coffee company deciding whether to launch a new drink.

Each day, you read six customer comments and selected the comments that counted as **genuine support** for launching the drink. The app recorded the number of selected comments as:

$$
x = \text{number of comments judged as genuine support}
$$

Each day had:

$$
n = 6
$$

We will use your copied data from the app to reproduce the Bayesian updating process in R.

The unknown parameter is:

$$
\theta = \text{the proportion of customers who genuinely support launching the drink}
$$

We begin with a neutral prior:

$$
\theta \sim \text{Beta}(1,1)
$$

After observing \(x\) supportive comments out of \(n\), the posterior is:

$$
\theta \mid x
\sim
\text{Beta}(\alpha + x,\ \beta + n - x)
$$

The posterior from one day becomes the prior for the next day.

```{r}
library(tidyverse)
library(gt)
```

:::: callout-note
#### Question 1

Paste the data copied from the app into your Quarto document.

Your copied data should look something like this:

```{r, eval=FALSE}
my_data <- c(4, 3, 5, 2, 4, 5)
```

Each value is the number of customer comments you selected as genuine support on that day.

```{r, eval=FALSE}
my_data <- c(___, ___, ___, ___, ___, ___)

my_data
```

::: {.callout-important collapse="true"}
#### Click for solution

Your values will be different from the example below, depending on how you classified the customer comments.

```{r}
my_data <- c(4, 3, 5, 2, 4, 5)

my_data
```

There should be six numbers because the app runs for six days.
:::
::::

:::: callout-note
#### Question 2

Create a tibble called `launch_data` with one row per day.

The variable `x` should contain the values from `my_data`. The variable `n` should be 6 for every day.

```{r, eval=FALSE}
launch_data <- tibble(
  day = ___,
  x = ___,
  n = ___
)

launch_data
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r}
launch_data <- tibble(
  day = 1:6,
  x = my_data,
  n = 6
)

launch_data
```

The variable `x` records the number of comments judged as genuine support. The variable `n` records the number of comments shown each day.
:::
::::

:::: callout-note
#### Question 3

Calculate the total number of supportive comments and the total number of comments reviewed.

Then calculate the overall sample proportion.

```{r, eval=FALSE}
launch_summary <- launch_data |>
  summarise(
    total_support = sum(___),
    total_comments = sum(___),
    sample_proportion = ___ / ___
  )

launch_summary
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r}
launch_summary <- launch_data |>
  summarise(
    total_support = sum(x),
    total_comments = sum(n),
    sample_proportion = total_support / total_comments
  )

launch_summary
```

This is the raw proportion of comments you classified as genuine support across all six days.
:::
::::

:::: callout-note
#### Question 4

Use a loop to update the beta distribution one day at a time.

Start with:

$$
\alpha = 1
$$

and:

$$
\beta = 1
$$

For each day, update using:

$$
\alpha_{new} = \alpha_{old} + x
$$

and:

$$
\beta_{new} = \beta_{old} + n - x
$$

```{r, eval=FALSE}
alpha_current <- ___
beta_current <- ___

updates <- list()

for (i in 1:nrow(launch_data)) {
  x_today <- launch_data$___[i]
  n_today <- launch_data$___[i]
  
  alpha_current <- alpha_current + ___
  beta_current <- beta_current + ___ - ___
  
  updates[[i]] <- tibble(
    day = launch_data$day[i],
    x = x_today,
    n = n_today,
    alpha = alpha_current,
    beta = beta_current,
    posterior_mean = alpha_current / (alpha_current + beta_current)
  )
}

posterior_updates <- bind_rows(updates)

posterior_updates
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r}
alpha_current <- 1
beta_current <- 1

updates <- list()

for (i in 1:nrow(launch_data)) {
  x_today <- launch_data$x[i]
  n_today <- launch_data$n[i]
  
  alpha_current <- alpha_current + x_today
  beta_current <- beta_current + n_today - x_today
  
  updates[[i]] <- tibble(
    day = launch_data$day[i],
    x = x_today,
    n = n_today,
    alpha = alpha_current,
    beta = beta_current,
    posterior_mean = alpha_current / (alpha_current + beta_current)
  )
}

posterior_updates <- bind_rows(updates)

posterior_updates
```

Each row gives the posterior distribution after that day of customer feedback.
:::
::::

:::: callout-note
#### Question 5

Add the initial prior to the update table and create a clean summary table.

The initial prior is:

$$
\text{Beta}(1,1)
$$

with mean:

$$
\frac{1}{1+1} = 0.50
$$

```{r, eval=FALSE}
initial_prior <- tibble(
  day = ___,
  x = ___,
  n = ___,
  alpha = ___,
  beta = ___,
  posterior_mean = ___
)

posterior_path <- bind_rows(
  ___,
  ___
) |>
  mutate(
    distribution = paste0("Beta(", ___, ", ", ___, ")")
  )

posterior_path |>
  select(
    Day = day,
    Support = x,
    Comments = n,
    Distribution = distribution,
    `Posterior mean` = posterior_mean
  )
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r}
initial_prior <- tibble(
  day = 0,
  x = 0,
  n = 0,
  alpha = 1,
  beta = 1,
  posterior_mean = 0.5
)

posterior_path <- bind_rows(
  initial_prior,
  posterior_updates
) |>
  mutate(
    distribution = paste0("Beta(", alpha, ", ", beta, ")")
  )

posterior_path |>
  dplyr::select(
    Day = day,
    Support = x,
    Comments = n,
    Distribution = distribution,
    `Posterior mean` = posterior_mean
  ) |>
  gt() |>
  fmt_number(columns = `Posterior mean`, decimals = 3)
```

This table shows how the beta distribution changes as more customer feedback is included.
:::
::::

:::: callout-note
#### Question 6

Plot the prior and posterior distributions across the six days.

First, create a grid of possible values for \(\theta\). Then calculate the beta density for each distribution in `posterior_path`.

```{r, eval=FALSE}
plot_data <- posterior_path |>
  mutate(
    label = if_else(day == ___, "Initial prior", paste0("Day ", day))
  ) |>
  crossing(theta = seq(0.001, 0.999, length.out = 500)) |>
  mutate(
    density = dbeta(theta, shape1 = ___, shape2 = ___),
    label = factor(label, levels = unique(label))
  )

ggplot(plot_data, aes(x = ___, y = ___)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ label, ncol = 1, scales = "free_y") +
  labs(
    x = expression(theta),
    y = "Density",
    title = "Prior and posterior distributions",
    subtitle = "Each panel shows belief after another day of feedback"
  ) +
  theme_minimal()
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r}
plot_data <- posterior_path |>
  mutate(
    label = if_else(day == 0, "Initial prior", paste0("Day ", day))
  ) |>
  tidyr::crossing(theta = seq(0.001, 0.999, length.out = 500)) |>
  mutate(
    density = dbeta(theta, shape1 = alpha, shape2 = beta),
    label = factor(label, levels = unique(label))
  )

ggplot(plot_data, aes(x = theta, y = density)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ label, ncol = 1, scales = "free_y") +
  labs(
    x = expression(theta),
    y = "Density",
    title = "Prior and posterior distributions",
    subtitle = "Each panel shows belief after another day of feedback"
  ) +
  theme_minimal()
```

The posterior should usually become more concentrated as more comments are classified. This happens because the model has more information after each day.
:::
::::

:::: callout-note
#### Question 7

Summarise the final posterior distribution.

Calculate:

1. the final posterior mean;
2. a 95% credible interval;
3. the posterior probability that more than half of customers genuinely support launching the drink.

```{r, eval=FALSE}
final_posterior <- posterior_path |>
  slice_tail(n = ___)

final_summary <- tibble(
  final_distribution = paste0("Beta(", final_posterior$___, ", ", final_posterior$___, ")"),
  posterior_mean = final_posterior$___,
  lower_95 = qbeta(___, shape1 = final_posterior$___, shape2 = final_posterior$___),
  upper_95 = qbeta(___, shape1 = final_posterior$___, shape2 = final_posterior$___),
  prob_above_half = 1 - pbeta(___, shape1 = final_posterior$___, shape2 = final_posterior$___)
)

final_summary
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r}
final_posterior <- posterior_path |>
  slice_tail(n = 1)

final_summary <- tibble(
  final_distribution = paste0("Beta(", final_posterior$alpha, ", ", final_posterior$beta, ")"),
  posterior_mean = final_posterior$posterior_mean,
  lower_95 = qbeta(0.025, shape1 = final_posterior$alpha, shape2 = final_posterior$beta),
  upper_95 = qbeta(0.975, shape1 = final_posterior$alpha, shape2 = final_posterior$beta),
  prob_above_half = 1 - pbeta(0.50, shape1 = final_posterior$alpha, shape2 = final_posterior$beta)
)

final_summary |>
  gt() |>
  fmt_number(columns = c(posterior_mean, lower_95, upper_95, prob_above_half), decimals = 3)
```

The value `prob_above_half` answers the question:

> After six days of customer feedback, how likely is it that more than half of customers genuinely support launching the drink?

A value close to 1 provides strong evidence in favour of launch support being above 50%. A value close to 0 suggests weak evidence that support is above 50%. A value near 0.5 means there is still substantial uncertainty.
:::
::::