---
title: "Week 2"
format: html
subtitle: "Statistical Thinking"
---

```{r}
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tinytex)
library(tidyverse)
```

## Introduction

This week we begin with a short witness activity.

You will watch a brief animated scene only once. Your task is to pay attention to what happens and then answer a series of recall questions. Some questions will be about the main event. Others will be about background details, such as objects in the room, people nearby, or things that were easy to overlook.

After the scene, you will answer each recall question and give a confidence rating. This gives us two kinds of data:

- whether your answer was correct
- how confident you were in that answer

This lets us ask better statistical questions. Instead of only asking:

> What happened?

we can ask:

> How reliable was the witness evidence?

> Did confidence match accuracy?

> Which details were easiest or hardest to remember?

> How much variation was there across witnesses?

```{=html}
<iframe
  src="interactive/witness.html"
  width="100%"
  height="640"
  style="border: none;"
></iframe>
```

## Tasks

### Part 1: Your witness data

In the app, `my_answers` records whether each recall answer was correct.

- `1` means your answer was correct
- `0` means your answer was incorrect

The vector `my_confidence` records how confident you were for each answer, on a 0 to 100 scale.

:::: callout-note
#### Question 1

Click the **Copy data for RStudio** button on the app above and paste your results into your Quarto document. It should look something like: 


```{r, eval=FALSE}
my_answers <- c(___, ___, ___, ___, ___, ___, ___, ___)
my_confidence <- c(___, ___, ___, ___, ___, ___, ___, ___)

```

::::

:::: callout-note
#### Question 2

Create a tibble called `witness_data` with one row for each recall question.

```{r, eval=FALSE}
witness_data <- tibble(
  question = ___,
  correct = ___,
  confidence = ___
)

witness_data
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r, eval=FALSE}
witness_data <- tibble(
  question = 1:length(my_answers),
  correct = my_answers,
  confidence = my_confidence
)

```

Each row represents one recall question from the witness activity.
:::
::::

### Part 2: Accuracy and confidence

A natural first question is:

> How many questions did I get right?

But a better statistical question is:

> How accurate was my recall, and how confident was I in that recall?

:::: callout-note
#### Question 3

Calculate your recall score and your recall accuracy.

```{r, eval=FALSE}
recall_summary <- witness_data |>
  summarise(
    number_correct = sum(___),
    number_questions = n(),
    accuracy = mean(___)
  )

```

::: {.callout-important collapse="true"}
#### Click for solution

```{r, eval=FALSE}
recall_summary <- witness_data |>
  summarise(
    number_correct = sum(correct),
    number_questions = n(),
    accuracy = mean(correct)
  )

recall_summary
```


:::
::::

:::: callout-note
#### Question 4

Calculate your average confidence.

```{r, eval=FALSE}
confidence_summary <- witness_data |>
  summarise(
    average_confidence = mean(___),
    lowest_confidence = min(___),
    highest_confidence = max(___)
  )

```

::: {.callout-important collapse="true"}
#### Click for solution

```{r, eval=FALSE}
confidence_summary <- witness_data |>
  summarise(
    average_confidence = mean(confidence),
    lowest_confidence = min(confidence),
    highest_confidence = max(confidence)
  )

```

Confidence is not the same as accuracy. A witness can be confident and wrong, or uncertain and correct.
:::
::::

### Part 3: Did confidence match accuracy?

One useful way to think about witness evidence is to compare confidence with accuracy.

If someone gives a high-confidence answer, we might be tempted to treat that answer as strong evidence. But in real life, confidence and accuracy do not always match perfectly.

:::: callout-note
#### Question 5

Create a new variable called `confidence_group`.

Use the following categories:

- `Low` for confidence below 50
- `Medium` for confidence from 50 to 79
- `High` for confidence 80 or above

```{r, eval=FALSE}
witness_data <- witness_data |>
  mutate(
    confidence_group = case_when(
      confidence < ___ ~ "Low",
      confidence < ___ ~ "Medium",
      TRUE ~ "High"
    )
  )

```

::: {.callout-important collapse="true"}
#### Click for solution

```{r, eval=FALSE}
witness_data <- witness_data |>
  mutate(
    confidence_group = case_when(
      confidence < 50 ~ "Low",
      confidence < 80 ~ "Medium",
      TRUE ~ "High"
    )
  )

witness_data
```
:::
::::

:::: callout-note
#### Question 6

For each confidence group, calculate the average accuracy.

```{r, eval=FALSE}
confidence_accuracy <- witness_data |>
  group_by(___) |>
  summarise(
    n_answers = n(),
    accuracy = mean(___),
    average_confidence = mean(___),
    .groups = "drop"
  )

confidence_accuracy
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r, eval=FALSE}
confidence_accuracy <- witness_data |>
  group_by(confidence_group) |>
  summarise(
    n_answers = n(),
    accuracy = mean(correct),
    average_confidence = mean(confidence),
    .groups = "drop"
  )

confidence_accuracy
```

This table helps us ask whether higher confidence was associated with higher accuracy for your recall answers.
:::
::::

:::: callout-note
#### Question 7

Create a plot showing confidence and accuracy for each question.

```{r, eval=FALSE}
ggplot(witness_data, aes(x = question, y = ___)) +
  geom_col() +
  geom_point(
    aes(y = ___ / 100),
    size = 3
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent
  ) +
  labs(
    x = "____",
    y = "____",
    title = "____"
  ) +
  theme_minimal()
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r, eval=FALSE}
ggplot(witness_data, aes(x = question, y = correct)) +
  geom_col() +
  geom_point(
    aes(y = confidence / 100),
    size = 3
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent
  ) +
  labs(
    x = "Recall question",
    y = "Accuracy and confidence",
    title = "Recall accuracy and confidence",
    subtitle = "Bars show correctness; points show confidence"
  ) +
  theme_minimal()
```

A bar at 100% means the answer was correct. A bar at 0% means the answer was incorrect. The point shows how confident you were in that answer.
:::
::::

## Homework

### Scenario: Dr.Arbuthnot’s baptism records

John Arbuthnot was a Scottish physician. In 1710 he published a paper[^week02-1] that examined birth records in London for each of the 82 years from 1629 to 1710. Actually the data he used were counts of baptisms, by year and gender, but these are used as a proxy for the number of births.

[^week02-1]: Arbuthnot, J. (1710). “An argument for divine providence, taken from the constant regularity observed in the births of both sexes,” \textit{Philosophical Transactions of the Royal Society}, Vol 27, pp 186–190.

### Data

The data from Arbuthnot’s study is contained in the file `arbuthnot.csv`. It consists of three columns, named `year`, `boys` and `girls`. The `boys` and `girls` columns show the number of each that were born in London for each year as specified in the `year` column.

<center>
```{r, echo=FALSE}
library(downloadthis)
download_file(
  path = "data/arbuthnot.csv",
  button_label = "arbuthnot",
  button_type = 'success'
)
```
</center><br>

Download and transfer the file `arbuthnot.csv` to your Project folder > week 2 > data

### Tasks

1.  What line of R code would you use to extract the vector containing just the counts of girls baptised? Would it be a good idea to print it in an assignment?

2. Using `ggplot()`, create a plot to visualise the number of girls baptised per year. Is there an apparent trend in the number of girls baptized over the years? How would you describe it? Are there any interesting features?

3. Create three new columns in your data:
    - `total`: the sum of boy baptisims and girl baptisms
    - `ratio`: the number of girl baptisms divided by the number of boy baptisms
    - `prop_girls`: the proportion of girl baptisms

4. Draw a line plot of the total number of baptisms per year. How often were there more boys born than girls?

5.  Draw a plot of the proportion of boys born over time. What do you see? Discuss this plot. Can it be improved?

6.  Use the internet or help in R to create a **TERRIBLE** graph, which has none of the good attributes from the previous question. Comment on what makes it terrible.


