---
title: "Week 10"
Subtitle: "MCMC"
---

```{r}
#| include: false
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE)
library(tinytex)
library(tidyverse)
library(gt)
library(scales)
library(MCMCpack)
```

## Introduction

This week we will begin by playing a short sampling game.

-   The screen shows a hidden landscape, covered in fog.
-   At each step, you propose a move. You are shown the height of the landscape (the density) at your current position and at the proposed position.
-   You then decide whether to **move** or **stay**.
-   Every position you occupy, including repeats when you stay, counts as one of your samples.
-   Your score depends on how well your 30 samples reproduce the true shape of the hidden landscape, not on how high up it you climb.

There are two rounds, each with a different hidden landscape. You also get 3 probes per round, which reveal the exact height of the landscape wherever you click, at a small cost to your score.

```{=html}
<iframe
  src="interactive/MCMC.html"
  width="100%"
  height="400"
  style="border: none;"
></iframe>
```

Play both rounds. Record your final score.

Once you have played the game, discuss with your tutorial group:

-   Did always moving toward higher density give you a good score? Why or why not?
-   In round 2, did you find the smaller hump? What helped you find it?
-   What happened to your histogram in the first few steps, before it had time to "settle"?

This is exactly what a Metropolis sampler does. It proposes a move, compares the (unnormalised) density at the current and proposed points, and moves with a probability based on that ratio. Run for long enough, the collection of visited points looks like a sample from the target distribution, even though the sampler never needed to know the total area under the curve.

<br>

## Tasks

**Note**: Create a Quarto document in your Project folder and save it in your `week 10` folder.

### Part 1: A hand-built sampler for a proportion

Suppose we want the posterior distribution for the probability of heads, $p$, after observing 12 heads in 20 coin flips, using a Uniform(0, 1) prior.

The posterior is proportional to the likelihood, since the prior is flat:

$$
\pi(p \mid \text{data}) \propto p^{12}(1-p)^{8}.
$$

`MCMCpack::MCMCmetrop1R()` will draw from this posterior for us. It needs a function that returns the **log** of the (unnormalised) posterior density, because working on the log scale avoids numerical underflow when densities get very small.

:::: callout-note
#### Question 1

Write the log-posterior function for $p$. Remember to return `-Inf` outside the support of $p$, so the sampler never accepts a move to an impossible value.

Fill in the blanks.

```{r, eval=FALSE}
log_post_coin <- function(p) {
  if (p <= 0 || p >= 1) return(___)
  dbinom(___, size = ___, prob = ___, log = ___)
}
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r}
log_post_coin <- function(p) {
  if (p <= 0 || p >= 1) return(-Inf)
  dbinom(12, size = 20, prob = p, log = TRUE)
}
```

Because the prior is Uniform(0, 1), it contributes a constant (1, or 0 on the log scale) over the support of $p$, so the log-likelihood alone is a valid log-posterior up to a constant. We still need the boundary check, since `dbinom` is only meaningful for `p` between 0 and 1.
:::
::::

:::: callout-note
#### Question 2

Use `MCMCmetrop1R()` to draw 10,000 samples from the posterior, after a burn-in of 1,000. Start the chain at $p = 0.5$.

Fill in the blanks.

```{r, eval=FALSE}
coin_chain <- MCMCmetrop1R(
  ___,
  theta.init = ___,
  mcmc = ___,
  burnin = ___,
  seed = 1
)

summary(coin_chain)
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r}
coin_chain <- MCMCmetrop1R(
  log_post_coin,
  theta.init = 0.5,
  mcmc = 10000,
  burnin = 1000,
  seed = 1
)

summary(coin_chain)
```

Two things are printed to the console when this runs. First, a report of the Metropolis acceptance rate. Second, the usual `summary()` output for an `mcmc` object: the posterior mean, standard deviation, and a set of quantiles.

The posterior mean should be close to $12/20 = 0.60$, pulled slightly toward $0.5$ by the small amount of information in the prior relative to a sample of size 20. In fact, since the prior is conjugate here, the exact posterior is $\text{Beta}(13, 9)$, with mean $13/22 \approx 0.591$ and standard deviation $\approx 0.103$. Your simulated summary should be close to these exact values.
:::
::::

:::: callout-note
#### Question 3

`MCMCmetrop1R()` reported an acceptance rate. Extract the chain as a plain vector and calculate this rate directly, as the proportion of iterations where the value changed from the previous one.

Fill in the blanks.

```{r, eval=FALSE}
coin_vec <- as.numeric(___)
acceptance_rate <- mean(diff(___) != 0)
acceptance_rate
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r}
coin_vec <- as.numeric(coin_chain)
acceptance_rate <- mean(diff(coin_vec) != 0)
acceptance_rate
```

This should match the rate reported by `MCMCmetrop1R()` at around 0.71, or 71%. A rate in the rough range of 20% to 50% is often considered efficient for random-walk Metropolis, though for a single simple parameter like this one, a higher rate is not a problem.
:::
::::

:::: callout-note
#### Question 4

Use the `summary()` function to explore your MCMC output

```{r, eval=FALSE}
summary(___)
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r}
summary(coin_chain)
```

:::
::::

:::: callout-note
#### Question 5

We'll create a dataset of the output so that we can create some plots. Have two columns `iteration` and `theta`. These should be based on the output above.

```{r, eval=FALSE}
chain_coin_df <- tibble(
  iteration = 1:length(___),
  theta = as.numeric(___)
)
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r, eval=FALSE}
chain_coin_df <- tibble(
  iteration = 1:length(coin_chain),
  theta = as.numeric(coin_chain)
)
```

:::
::::

:::: callout-note
#### Question 6

Using the `chain_coin_df` from the previous question, create a trace plot.

```{r, eval=FALSE}
p_trace <- ggplot(___, aes(x = ___, y = ___)) +
  geom_point(size = 0.1) +
  geom_line()
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r, eval=FALSE}
p_trace <- ggplot(chain_coin_df, aes(x = iteration, y = theta)) +
  geom_point(size = 0.1) +
  geom_line()
```

:::
::::

:::: callout-note
#### Question 7

Create a density plot of the data

```{r, eval=FALSE}
p_density <- ggplot(___, aes(x = ___)) +
  geom___
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r, eval=FALSE}
p_density <- ggplot(chain_coin_df, aes(x = theta)) +
  geom_density()
```

:::
::::

:::: callout-note
#### Question 8

Use `patchwork` to place the two plots together

```{r, eval=FALSE}
library(patchwork)

___ + ___
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r, eval=FALSE}
p_trace + p_density
```

:::
::::

### Part 2: A two-humped posterior, and checking your chain

Round 2 of the warm-up game had a hidden landscape with two humps: a large one and a smaller, easy-to-miss one. This is exactly the shape you get when a parameter has two plausible explanations behind it. Suppose $\theta$, between 0 and 10, represents a quantity that could plausibly come from either of two subgroups: a large subgroup centred around 3, and a smaller subgroup centred around 7.5. As an unnormalised posterior, this is a two-component mixture:

$$
\pi(\theta \mid \text{data}) \propto 0.72 \times \text{Normal}(\theta; 3, 0.6) + 0.28 \times \text{Normal}(\theta; 7.5, 0.5).
$$

:::: callout-note
#### Question 9

Write the log-posterior function for $\theta$, and draw 10,000 samples using `MCMCmetrop1R()`, with a burn-in of 1,000 and `theta.init = 1`, matching the starting position in round 2 of the game.

Fill in the blanks.

```{r, eval=FALSE}
log_post_two_groups <- function(theta) {
  if (theta <= 0 || theta >= 10) return(___)
  log(___ * dnorm(theta, mean = ___, sd = ___) + ___ * dnorm(theta, mean = ___, sd = ___))
}

two_groups_chain <- MCMCmetrop1R(
  ___,
  theta.init = ___,
  mcmc = 10000,
  burnin = 1000,
  seed = 1
)
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r}
log_post_two_groups <- function(theta) {
  if (theta <= 0 || theta >= 10) return(-Inf)
  log(0.72 * dnorm(theta, mean = 3, sd = 0.6) + 0.28 * dnorm(theta, mean = 7.5, sd = 0.5))
}

two_groups_chain <- MCMCmetrop1R(
  log_post_two_groups,
  theta.init = 1,
  mcmc = 10000,
  burnin = 1000,
  seed = 1
)
```

:::
::::

:::: callout-note
#### Question 10

Use the `summary()` function to explore your MCMC output.

```{r, eval=FALSE}
summary(___)
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r}
summary(two_groups_chain)
```

:::
::::

:::: callout-note
#### Question 11

We'll create a dataset of the output so that we can create some plots. Have two columns `iteration` and `theta`. These should be based on the output above.

```{r, eval=FALSE}
chain_two_groups_df <- tibble(
  iteration = 1:length(___),
  theta = as.numeric(___)
)
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r}
chain_two_groups_df <- tibble(
  iteration = 1:length(two_groups_chain),
  theta = as.numeric(two_groups_chain)
)
```
:::
::::

:::: callout-note
#### Question 12

Using the `chain_two_groups_df` from the previous question, create a trace plot.

```{r, eval=FALSE}
p_trace <- ggplot(___, aes(x = ___, y = ___)) +
  geom_point(size = 0.1) +
  geom_line()
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r}
p_trace <- ggplot(chain_two_groups_df, aes(x = iteration, y = theta)) +
  geom_point(size = 0.1) +
  geom_line()

p_trace
```

Look for the chain jumping back and forth between a region near 3 and a region near 7.5. That jumping is the sampler moving between the two humps. If your chain only ever visited one hump, that is the multimodal equivalent of a student in the game who never found the second hill: the sampler is not wrong, it just has not been given the chance (or the step size) to discover the other mode.
:::
::::

:::: callout-note
#### Question 13

Create a density plot of the data.

```{r, eval=FALSE}
p_density <- ggplot(___, aes(x = ___)) +
  geom___
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r}
p_density <- ggplot(chain_two_groups_df, aes(x = theta)) +
  geom_density()

p_density
```

The density plot should show two bumps, a tall one near 3 and a shorter one near 7.5, mirroring the shape of the fog-covered landscape from round 2 of the warm-up game, now filled in by your samples instead of by probes.
:::
::::

:::: callout-note
#### Question 14

Use `patchwork` to place the two plots together.

```{r, eval=FALSE}
library(patchwork)

___ + ___
```

::: {.callout-important collapse="true"}
#### Click for solution

```{r}
library(patchwork)

p_trace + p_density
```

:::
::::


















