---
title: "Week 8"
format: html
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE)
library(tinytex)
library(tidyverse)
```

In this tutorial we will begin by investigating **likelihood** and **fitting models** using the app below. Your tutor will explain to you how to use it.

```{=html}
<div style="width: 100%; height: 760px;">
  <iframe
    src="interactive/model-speed-dating.html"
    title="MLE Matchmaker"
    style="width: 100%; height: 100%; border: none; border-radius: 12px;"
    loading="lazy"
    allowfullscreen>
  </iframe>
</div>
```

# Introduction

In the opening activity, you selected probability distributions and adjusted their parameters until they appeared to fit the observed data. You were informally searching for the combination of model and parameter values that made the data most likely.

In this tutorial, R will perform that search for us. We will use:

- `MASS::fitdistr()` to obtain maximum likelihood estimates;
- the fitted log-likelihood to compare candidate models; and
- Q–Q plots to investigate how well each fitted distribution describes the data.

# Scenario

A hospital is investigating patient activity in its emergency department. For each of 200 patients, two variables were recorded:

- `waiting_time`: the number of minutes before the patient was seen;
- `tests_ordered`: the number of diagnostic tests ordered for the patient.

The hospital wants to identify probability models that adequately describe these variables. Appropriate models could later be used for simulation, planning and estimating the probability of unusually long waits or unusually large numbers of tests.

Download the data below:

<center>

```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
library(downloadthis)

download_file(
  path = "data/emergency.csv",
  button_label = "emergency",
  button_type = 'success'
)
```

</center>

<br>

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
emergency_data <- read.csv("data/emergency.csv")
```


# Tasks

:::: callout-note
## Question 1: EDA

Use `glimpse()` and `summary()` to inspect `emergency_data`. Then create:

1. a histogram of `waiting_time`; and
2. a bar chart of `tests_ordered`.

What features of each variable will be important when choosing a probability model?

```{r, eval=FALSE}
glimpse(___)
summary(___)
```

```{r, eval=FALSE}
# Note: adding after_stat() allows us to visualise the densities
ggplot(emergency_data, aes(x = ___)) +
  geom_histogram(aes(y = after_stat(density)), bins = 25)
```

```{r, eval=FALSE}
ggplot(emergency_data, aes(x = factor(___))) +
  geom_bar()
```


::: {.callout-important collapse="true"}
## Click for solution

```{r}
glimpse(emergency_data)
summary(emergency_data)
```

```{r}
ggplot(emergency_data, aes(x = waiting_time)) +
  geom_histogram(
    aes(y = after_stat(density))
  )
```

```{r}
ggplot(emergency_data, aes(x = factor(tests_ordered))) +
  geom_bar()
```


`waiting_time` is positive, continuous and right-skewed. `tests_ordered` contains non-negative integer counts and is also right-skewed. These features help determine which distributions are plausible.
:::
::::

:::: callout-note
## Question 2: Fit models to waiting times

Fit a normal distribution to `waiting_time` using `fitdistr()`. Save the result as `waiting_normal` and print it.

```{r, eval=FALSE}
library(MASS)
waiting_normal <- fitdistr(
  emergency_data$___,
  densfun = "___"
)

waiting_normal
```

Identify the maximum likelihood estimates of the mean and standard deviation.

::: {.callout-important collapse="true"}
## Click for solution

```{r}
library(MASS)
waiting_normal <- fitdistr(
  emergency_data$waiting_time,
  densfun = "normal"
)

waiting_normal
```

The values under `mean` and `sd` are the maximum likelihood estimates. They are the parameter values that maximise the normal model's likelihood for the observed waiting times.
:::
::::

:::: callout-note
## Question 3

Fit exponential, gamma, Weibull and log-normal distributions to `waiting_time`.

```{r, eval=FALSE}
waiting_exponential <- fitdistr(
  emergency_data$waiting_time,
  densfun = "___"
)

waiting_gamma <- fitdistr(
  emergency_data$waiting_time,
  densfun = "___"
)

waiting_weibull <- fitdistr(
  emergency_data$waiting_time,
  densfun = "___"
)

waiting_lognormal <- fitdistr(
  emergency_data$waiting_time,
  densfun = "___"
)
```

Print each fitted model and identify its maximum likelihood estimates.

::: {.callout-important collapse="true"}
## Click for solution

```{r}
waiting_exponential <- fitdistr(
  emergency_data$waiting_time,
  densfun = "exponential"
)

waiting_gamma <- fitdistr(
  emergency_data$waiting_time,
  densfun = "gamma"
)

waiting_weibull <- fitdistr(
  emergency_data$waiting_time,
  densfun = "weibull"
)

waiting_lognormal <- fitdistr(
  emergency_data$waiting_time,
  densfun = "lognormal"
)

waiting_exponential
waiting_gamma
waiting_weibull
waiting_lognormal
```

`fitdistr()` automatically selects the parameter values that maximise the likelihood for each candidate distribution.
:::
::::

:::: callout-note
## Question 4: Compare fitted log-likelihoods

The fitted log-likelihood can be extracted using `$loglik`.

Complete the code to create a comparison table.

```{r}
waiting_comparison <- tibble(
  model = c(
    "Normal",
    "Exponential",
    "Gamma",
    "Weibull",
    "Log-normal"
  ),
  log_likelihood = c(
    waiting_normal$___,
    waiting_exponential$___,
    waiting_gamma$___,
    waiting_weibull$___,
    waiting_lognormal$___
  )
) |>
  arrange(desc(log_likelihood))

waiting_comparison
```

Which fitted model has the largest log-likelihood?

::: {.callout-important collapse="true"}
## Click for solution

```{r}
waiting_comparison <- tibble(
  model = c(
    "Normal",
    "Exponential",
    "Gamma",
    "Weibull",
    "Log-normal"
  ),
  log_likelihood = c(
    waiting_normal$loglik,
    waiting_exponential$loglik,
    waiting_gamma$loglik,
    waiting_weibull$loglik,
    waiting_lognormal$loglik
  )
) |>
  arrange(desc(log_likelihood))

waiting_comparison
```

The model at the top of the table has the largest maximised log-likelihood. A larger value means that the fitted model assigns greater likelihood to the observed data.

The log-likelihood should not be the only evidence used to assess a model. We will also inspect the fitted shape and Q–Q plots.
:::
::::

:::: callout-note
## Question 5

`ggplot2` provides two functions for constructing Q–Q plots:

- `stat_qq()` adds the observed points;
- `stat_qq_line()` adds the reference line.

For distributions other than the standard normal distribution, we provide:

- the relevant quantile function using `distribution`; and
- the fitted parameters using `dparams = as.list(fit$estimate)`.

The required quantile functions are:

| Distribution | Quantile function |
|---|---|
| Normal | `qnorm` |
| Exponential | `qexp` |
| Gamma | `qgamma` |
| Weibull | `qweibull` |
| Log-normal | `qlnorm` |
| Poisson | `qpois` |
| Negative binomial | `qnbinom` |

Create Q–Q plots for the fitted normal and gamma models.

```{r, eval=FALSE}
ggplot(emergency_data, aes(sample = waiting_time)) +
  stat_qq(
    distribution = ___,
    dparams = as.list(___$estimate)
  ) +
  stat_qq_line(
    distribution = ___,
    dparams = as.list(___$estimate)
  )

ggplot(emergency_data, aes(sample = waiting_time)) +
  stat_qq(
    distribution = ___,
    dparams = as.list(___$estimate)
  ) +
  stat_qq_line(
    distribution = ___,
    dparams = as.list(___$estimate)
  )
```

Which model produces points that follow the reference line more closely?

::: {.callout-important collapse="true"}
## Click for solution

```{r}
ggplot(emergency_data, aes(sample = waiting_time)) +
  stat_qq(
    distribution = qnorm,
    dparams = as.list(waiting_normal$estimate)
  ) +
  stat_qq_line(
    distribution = qnorm,
    dparams = as.list(waiting_normal$estimate)
  )

ggplot(emergency_data, aes(sample = waiting_time)) +
  stat_qq(
    distribution = qgamma,
    dparams = as.list(waiting_gamma$estimate)
  ) +
  stat_qq_line(
    distribution = qgamma,
    dparams = as.list(waiting_gamma$estimate)
  )
```

The gamma Q–Q plot should follow the reference line more closely. Systematic curvature in the normal Q–Q plot indicates that the normal distribution does not adequately capture the right-skewed waiting times.
:::
::::

:::: callout-note
## Question 6

Create Q–Q plots for the exponential, gamma, Weibull and log-normal models. Use `patchwork` to display them in a two-by-two arrangement.

```{r}
library(patchwork)

qq_exponential <-
  ggplot(emergency_data, aes(sample = waiting_time)) +
  stat_qq(
    distribution = ___,
    dparams = as.list(___$estimate)
  ) +
  stat_qq_line(
    distribution = ___,
    dparams = as.list(___$estimate)
  ) +
  labs(title = "Exponential")

# Repeat for the gamma, Weibull and log-normal models.

qq_exponential + qq_gamma + qq_weibull + qq_lognormal
```

Which model would you select for `waiting_time`? Refer to both the log-likelihood comparison and Q–Q plots.

::: {.callout-important collapse="true"}
## Click for solution

```{r}
library(patchwork)

qq_exponential <-
  ggplot(emergency_data, aes(sample = waiting_time)) +
  stat_qq(
    distribution = qexp,
    dparams = as.list(waiting_exponential$estimate)
  ) +
  stat_qq_line(
    distribution = qexp,
    dparams = as.list(waiting_exponential$estimate)
  ) +
  labs(title = "Exponential")

qq_gamma <-
  ggplot(emergency_data, aes(sample = waiting_time)) +
  stat_qq(
    distribution = qgamma,
    dparams = as.list(waiting_gamma$estimate)
  ) +
  stat_qq_line(
    distribution = qgamma,
    dparams = as.list(waiting_gamma$estimate)
  ) +
  labs(title = "Gamma")

qq_weibull <-
  ggplot(emergency_data, aes(sample = waiting_time)) +
  stat_qq(
    distribution = qweibull,
    dparams = as.list(waiting_weibull$estimate)
  ) +
  stat_qq_line(
    distribution = qweibull,
    dparams = as.list(waiting_weibull$estimate)
  ) +
  labs(title = "Weibull")

qq_lognormal <-
  ggplot(emergency_data, aes(sample = waiting_time)) +
  stat_qq(
    distribution = qlnorm,
    dparams = as.list(waiting_lognormal$estimate)
  ) +
  stat_qq_line(
    distribution = qlnorm,
    dparams = as.list(waiting_lognormal$estimate)
  ) +
  labs(title = "Log-normal")

qq_exponential + qq_gamma + qq_weibull + qq_lognormal
```

Gamma should provide a strong overall fit. It should have one of the largest log-likelihoods and a Q–Q plot that remains close to the reference line. Weibull may also appear plausible, so conclusions should be based on the observed evidence rather than requiring a perfectly unambiguous answer.
:::
::::

# Homework

Repeat the above for the discrete variable `tests_ordered`.

:::: callout-note
## Question 7

Fit Poisson and negative binomial distributions to `tests_ordered`.

```{r}
tests_poisson <- fitdistr(
  emergency_data$tests_ordered,
  densfun = "___"
)

tests_negative_binomial <- fitdistr(
  emergency_data$tests_ordered,
  densfun = "___"
)

tests_poisson
tests_negative_binomial
```

Compare their fitted log-likelihoods. Which model is more strongly supported?

::: {.callout-important collapse="true"}
## Click for solution

```{r}
tests_poisson <- fitdistr(
  emergency_data$tests_ordered,
  densfun = "Poisson"
)

tests_negative_binomial <- fitdistr(
  emergency_data$tests_ordered,
  densfun = "negative binomial"
)

tests_poisson
tests_negative_binomial

tibble(
  model = c("Poisson", "Negative binomial"),
  log_likelihood = c(
    tests_poisson$loglik,
    tests_negative_binomial$loglik
  )
) |>
  arrange(desc(log_likelihood))
```

The negative binomial model should have the larger log-likelihood because the simulated counts contain more variability than a Poisson model can accommodate.
:::
::::

:::: callout-note
## Question 8

Create Q–Q plots for the Poisson and negative binomial models.

```{r}
qq_poisson <-
  ggplot(emergency_data, aes(sample = tests_ordered)) +
  stat_qq(
    distribution = ___,
    dparams = as.list(___$estimate)
  ) +
  stat_qq_line(
    distribution = ___,
    dparams = as.list(___$estimate)
  )

qq_negative_binomial <-
  ggplot(emergency_data, aes(sample = tests_ordered)) +
  stat_qq(
    distribution = ___,
    dparams = as.list(___$estimate)
  ) +
  stat_qq_line(
    distribution = ___,
    dparams = as.list(___$estimate)
  )

qq_poisson + qq_negative_binomial
```

Which model provides the better overall fit?

::: {.callout-important collapse="true"}
## Click for solution

```{r}
qq_poisson <-
  ggplot(emergency_data, aes(sample = tests_ordered)) +
  stat_qq(
    distribution = qpois,
    dparams = as.list(tests_poisson$estimate)
  ) +
  stat_qq_line(
    distribution = qpois,
    dparams = as.list(tests_poisson$estimate)
  )

qq_negative_binomial <-
  ggplot(emergency_data, aes(sample = tests_ordered)) +
  stat_qq(
    distribution = qnbinom,
    dparams = as.list(tests_negative_binomial$estimate)
  ) +
  stat_qq_line(
    distribution = qnbinom,
    dparams = as.list(tests_negative_binomial$estimate)
  )

qq_poisson + qq_negative_binomial
```

The negative binomial model should provide the better overall fit. Its log-likelihood is larger and its Q–Q plot should display less systematic departure.

Because count data contain repeated integer values, their Q–Q plots will naturally appear stepped. Look for the overall pattern rather than expecting every point to lie perfectly on the line.
:::
::::

:::: callout-note
## Question 9

In two or three sentences, recommend:

1. a probability model for patient waiting times; and
2. a probability model for the number of tests ordered.

Support each recommendation using the fitted log-likelihood and Q–Q plot.

::: {.callout-important collapse="true"}
## Click for solution

A gamma distribution is a suitable model for patient waiting times because it has a relatively large fitted log-likelihood and its Q–Q plot follows the reference line reasonably closely. A negative binomial distribution is suitable for the number of tests because it fits the overdispersed counts better than Poisson according to both its log-likelihood and Q–Q plot.

These models are the strongest choices among those considered; the diagnostics do not prove that either is the true data-generating distribution.
:::
::::






